---
title: "Solo Project: Internal Migration in Lithuania"
format: html
number-sections: false
---

# Process

```{r}
# load packages
library(tidyverse)
library(rnaturalearth)
library(RColorBrewer)
```

```{r}
# load data
lt_internal_migration <- read_csv("../data/net_migration_apskritys.csv")

head(lt_internal_migration)
```


```{r}
# load map
lt_apskritys <- ne_states(country = "Lithuania")
```

```{r}
# cleaning data
avg_internal_clean <- lt_internal_migration |>
  rename(
    year = Laikotarpis,
    region = `Administracinė teritorija`,
    migration = Reikšmė
  ) |> # translate relevant vars
  select(year, region, migration) |>
  filter(str_detect(region, "apskritis")) |> #removing irrelevant info
  mutate(region = str_remove(region, " apskritis")) |>
  mutate(region = fct_recode(region,
                             "Alytus" = "Alytaus",
                             "Kaunas" = "Kauno",
                             "Klaipėda" = "Klaipėdos",
                             "Marijampolė" = "Marijampolės",
                             "Panevėžys" = "Panevėžio",
                             "Tauragė" = "Tauragės",
                             "Telšiai" = "Telšių",
                             "Utena" = "Utenos",
                             "Vilnius" = "Vilniaus",
                             "Šiauliai" = "Šiaulių")) |> # compatibility edits
  group_by(region) |>
  summarise(avg_net_im = mean(migration)) |> # average net migration
  mutate(migration_cats = 
           cut(avg_net_im,
              breaks = c(-1527, -1000, -750, -450, 0, 1000, 2000, 4000),
              labels = c("-1527 to -1000", "-999 to -750", "-749 to -473", "-472 to 762", "763 to 1000", "1001 to 1500", "1501 to 3897"), # discretizing for better color control
              include.lowest = TRUE
  ))

avg_internal_clean
```

```{r}
# adding alt text and caption in next section
# create map of average net internal migration 2019-2023
lt_apskritys |>
  left_join(avg_internal_clean, join_by(name_en == region)) |>
  ggplot() +
  geom_sf(aes(fill = migration_cats)) +
  theme_void() +
  scale_fill_manual(values = brewer.pal(n = 6, name = "PRGn"), name = "# Migrants (net)") +
  labs(
    title = "Average Net Internal Migration in Lithuania, 2019-2023",
    caption = "Author: Quinn Cowell | Date: 4/20/2024 | Source: Valstybės duomenų agentūra, 2024")
```

## Final product

```{r}
#| echo: false
#| fig-cap: "Choropleth map of average annual net interal migration in Lithuania between 2019 to 2023, shown by net gains or losses of residents by county."
#| fig-alt: "Choropleth map of annual net interal migration in Lithuania from 2019 to 2023. Klaipėda, Vilnius, and Kaunas are the only counties with net gains; everywhere else is losing population."
# map
lt_apskritys |>
  left_join(avg_internal_clean, join_by(name_en == region)) |>
  ggplot() +
  geom_sf(aes(fill = migration_cats)) +
  theme_void() +
  scale_fill_manual(values = brewer.pal(n = 6, name = "PRGn"), name = "# Migrants (net)") +
  labs(
    title = "Average Net Internal Migration in Lithuania, 2019-2023",
    caption = "Author: Quinn Cowell | Date: 4/20/2024 | Source: Valstybės duomenų agentūra, 2024")
```

It is a well-known fact, at least in Lithuania, that the small Baltic state has been facing a decline in population since joining the EU in 2003, as Lithuanians seek better opportunities in the wealthier countries of Western Europe. However, a less widely-discussed migratory phenomenon is the movement of Lithuanians within their own country. I studied abroad in Vilnius, Lithuania in the fall of 2023. Not only is Vilnius a surprisingly international city, but also a destination city for Lithuanians with roots all across the country, from Alytus to Šilutė. This made me wonder: where do Lithuanians move to, and where do they move from? I gathered data from the Lithuanian state data agency and coded a map. The choropleth map above shows the average annual net internal migration in Lithuania within the period 2019-2023, divided by county (Lithuanian: apskritis) between the years; all counties share a name with their respective capital city. Green counties experienced a net gain of residents, while purple counties experienced a net loss. Unsurprisingly, the three green counties contain the three largest cities in Lithuania: Vilnius, the capital and largest city, experienced the greatest gain in residents by far, while Klaipėda, a seaside resort town, and Kaunas, in central Lithuania, experienced the second- and third-greatest gains in residents. Interestingly, the county containing the fourth-largest city in Lithuania, Šiauliai, experienced the greatest net decrease in population. This may be because, anecdotally, Šiauliai (as well as Lithuania's fifth-largest city, Panevėžys) has a reputation for being particularly economically depressed and crime-ridden. These are only stereotypes, but these two do top the list in terms of net population loss. These values may also be impacted by population differences between the counties, so there could be other counties that have worse outmigration crises than Šiauliai and Panevėžys which are obscured due to smaller population size.
