---
title: "Multivariate Viz"
---

## The story

Though far from a perfect assessment of academic preparedness, SAT scores have historically been used as one measurement of a state’s education system. The `education` dataset contains various education variables for each state:

```{r}
library(tidyverse)
# Import and check out data
education <- read.csv("https://mac-stat.github.io/data/sat.csv")
head(education)
```

## Exercise 1: SAT scores

```{r}
# Construct a plot of how the average sat scores vary across states (HINT: you need to use the sat variable only).
ggplot(education, aes(x = sat)) +
  geom_density()
```

### Exercise 2: SAT Scores vs Per Pupil Spending & SAT Scores vs Salaries

```{r}
# Construct a plot of sat vs expend
# Include a "best fit linear regression model" (HINT: method = "lm")
ggplot(education, aes(x = expend, y = sat)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
# Construct a plot of sat vs salary
# Include a "best fit linear regression model" (HINT: method = "lm")
ggplot(education, aes(x = salary, y = sat)) +
  geom_point() +
  geom_smooth(method = "lm")
```

# Exercise 3: SAT Scores vs Per Pupil Spending and Teacher Salaries

```{r}
# Construct one visualization of the relationship of sat with salary and expend. HINT: Start with just 2 variables and tweak that code to add the third variable. 
ggplot(education, aes(x = expend, y = sat, size = salary)) +
  geom_point()
```

## Exercise 4: Another way to Incorporate Scale

It can be tough to distinguish color scales and size scales for quantitative variables. Another option is to discretize a quantitative variable, or basically cut it up into categories.

Construct the plot below. Check out the code and think about what’s happening here. What happens if you change “2” to “3”?

```{r}
ggplot(education, aes(y = sat, x = salary, color = cut(expend, 2))) + 
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm")
```

## Exercise 5: Finally an Explanation

It’s strange that SAT scores seem to decrease with spending. But we’re leaving out an important variable from our analysis: the fraction of a state’s students that actually take the SAT. The fracCat variable indicates this fraction: low (under 15% take the SAT), medium (15-45% take the SAT), and high (at least 45% take the SAT).

### Part a

Build a univariate viz of fracCat to better understand how many states fall into each category.

```{r}
ggplot(education, aes(x = fracCat)) +
  geom_bar()
```


### Part b

Build 2 bivariate visualizations that demonstrate the relationship between sat and fracCat. What story does your graphic tell and why does this make contextual sense?

```{r}
ggplot(education, aes(x = sat, fill = fracCat)) +
  geom_density(alpha = .6)
```

```{r}
ggplot(education, aes(x = fracCat, y = sat)) +
  geom_violin()
```


### Part c

Make a trivariate visualization that demonstrates the relationship of sat with expend AND fracCat. Highlight the differences in fracCat groups through color AND unique trend lines. What story does your graphic tell?
Does it still seem that SAT scores decrease as spending increases?

```{r}
ggplot(education, aes(x = expend, y = sat, color = fracCat)) +
  geom_point() +
  geom_smooth()
```

